---
title: "Data records"
format: html
editor: visual
execute: 
  echo: false
  message: true
---

```{r}
#|label: set-up

library(targets)
library(tarchetypes)
library(tidyverse)
library(vegan)
library(nlme)
library(broom.mixed)
library(patchwork)

tar_load(community_itex_clean)
tar_load(community_gradient_clean)
tar_load(height_itex_clean)
tar_load(comm_structure_gradient_clean)
tar_load(traits_itex_clean)
tar_load(traits_gradient_clean)
tar_load(soil_gradient_clean)
tar_load(climate_gradient_clean)
tar_load(coords)
tar_load(temperature_itex_clean)
tar_load(climate_itex_clean)
```

```{r}
#|label: itex-comm

diversity <- community_itex_clean |> 
  group_by(Year, Site, Treatment, PlotID) |> 
  summarise(richness = n(),
            diversity = diversity(Abundance),
            evenness = diversity/log(richness),
            # proportions
            sum_abundance = sum(Abundance)) |> 
  pivot_longer(cols = c(richness:sum_abundance), names_to = "index", values_to = "value")

total <- diversity |> 
  group_by(index) |> 
  summarise(mean = mean(value),
            se = sd(value)/(sqrt(n())))

rich <- diversity |> 
  filter(index == "richness") |> 
  group_by(Site, Treatment) |> 
  summarise(mean = mean(value),
            se = sd(value)/(sqrt(n()))) |> 
  arrange(Treatment)

rich |> 
  select(-se) |> 
  pivot_wider(names_from = Treatment, values_from = mean) |> 
  mutate(diff = CTL - OTC)


```

## Data Records

### Plant community composition (dataset i)

The plot-level plant community dataset from the ITEX warming experiment has a total of `r community_itex_clean |> distinct(Taxon) |> count()` taxa and `r nrow(community_itex_clean)` observations (taxa x plots x years) (Table 1, Table 2). Overall mean species richness per plot, treatment and year (mean ± SE) is `r total$mean[3]` ± `r total$se[3]` species, including vascular plants, bryophytes and lichens. The species richness was `r rich$mean[3]` ± `r rich$se[3]` in the snowbed (SB), to `r rich$mean[1]` ± `r rich$se[1]` in the *Cassiope* heath and `r rich$mean[2]` ± `r rich$se[2]` in the *Dryas* heath (DH). Shannon diversity and evenness show the same pattern. The warming treatment had a very small effect on the community composition... For more details see().

```{r}
#|label: gradient-comm

diversity_grad <- community_gradient_clean |>
  group_by(Gradient, Site, PlotID) |>
  summarise(richness = n(),
            diversity = diversity(Cover),
            evenness = diversity/log(richness),
            # proportions
            sum_abundance = sum(Cover)) |>
  pivot_longer(cols = c(richness:sum_abundance), names_to = "index", values_to = "value")

total_grad <- diversity_grad |>
  group_by(index) |>
  summarise(mean = mean(value),
            se = sd(value)/(sqrt(n())))

rich_grad <- diversity_grad |>
  filter(index == "richness") |>
  group_by(Gradient, index) |>
  summarise(mean = mean(value),
            se = sd(value)/(sqrt(n())))


```

The plot-level plant community dataset from the gradients has a total of `r community_gradient_clean |> distinct(Taxon) |> count()` taxa and `r nrow(community_gradient_clean)` observations (taxa x plots x years) (Table 1, Table 3). 
Mean species richness (including graminoids, forbs and bryophytes) per plot is `r rich_grad$mean[2]` ± `r rich_grad$se[2]` species for the reference gradient and `r rich_grad$mean[1]` ± `r rich_grad$se[1]` for the nutrient gradient.
Shannon diversity and evenness were also slightly higher at the reference gradient.
For more details see bird cliff paper.

## Vegetation height and structure (dataset ii)

```{r}
#|label: itex-height

# test difference between treatment
height_itex_clean |> 
  group_by(Year, Site) |> 
  nest(data = -c(Year, Site)) |> 
  mutate(model = map(data, ~lme(height ~ Treatment, random = ~1|PlotID, data = .x)),
         result = map(model, tidy)) %>%
  unnest(result) |> 
  filter(effect == "fixed")
```

Vegetation height and structure data from the ITEX warming experiment has a total of `r nrow(height_itex_clean)` observations (site x treatment x plot; Table 1, Table 4). 
Vegetation did not differ between the control and warming treatment.

```{r}
#|label: gradient-height

height_grad <- comm_structure_gradient_clean |> 
  filter(Variable == "MedianHeight_cm") |> 
  group_by(Gradient) |> 
  summarise(mean = mean(Value),
            se = sd(Value/sqrt(n())))

comm_structure_gradient_clean |> 
  mutate(Gradient = factor(Gradient, levels = c("C", "B"))) |> 
  filter(Variable %in% c("Vascular", "MedianHeight_cm")) |> 
  group_by(Variable) |> 
  nest(data = -c(Variable)) |> 
  mutate(model = map(data, ~lme(Value ~ Gradient * Elevation_m, random = ~1|PlotID, data = .x)),
         result = map(model, tidy)) %>%
  unnest(result) |> 
  filter(effect == "fixed") |> 
  select(-data, -model, -effect, -group)

```

Vegetation height and structure data from the gradient has a total of `r round(nrow(comm_structure_gradient_clean), 2)` observations (gradient x site x plot x variable; Table 1, Table 5). 
Vegetation height increased with elevation at the bird cliff (E = 0.01 ± 0.004, t^1,80^ = 2.56, *P* = 0.013), but not at the reference site. 
The vascular plant cover decreased with elevation (E = -0.06 ± 0.026, t^1,80^ = -2.14, *P* = 0.035), but did not differ between the two gradients.



## Plant functional traits (dataset iii)

In the ITEX warming experiment, we measured physical, structural, chemical and nutrient traits (plant height, wet mass, dry mass, leaf area, leaf thickness, specific leaf area [SLA], and leaf dry matter content [LDMC], Carbon [C], Nitrogen [N], Phosphorus, C:N and NP ratios, and isotope [d13C, d15N]) for `r traits_itex_clean |> distinct(ID) |> count()` leaf samples from `r traits_itex_clean |> distinct(Taxon) |> count()` taxa across all sites and treatments, for a total of `r nrow(traits_itex_clean)` trait observations (site x treatment x plot; Table 1, Table 6). 
There are similar numbers of leaves per site (DH=1894; CH=1737; CB=1666) and treatment (CTL=2691; OTC=2606).

Along the gradients, we measured physical, structural, chemical and nutrient traits (plant height, wet mass, dry mass, leaf area, leaf thickness, shoot length, shoot ratio, specific leaf area [SLA], and leaf dry matter content [LDMC], water holding capacity [WHC], specific shoot length [SSL], Carbon [C], Nitrogen [N], Phosphorus, C:N and NP ratios, and isotope [d13C, d15N]) for `r traits_gradient_clean |> distinct(ID) |> count()` leaf samples from `r traits_gradient_clean |> distinct(Taxon) |> count()` taxa across all sites and treatments, for a total of `r nrow(traits_gradient_clean)` trait observations (site x treatment x plot; Table 1, Table 7).
Of those were `r traits_gradient_clean |> filter(Functional_group == "bryophyte") |> nrow()` observations and `r traits_gradient_clean |> filter(Functional_group == "bryophyte") |> distinct(Taxon) |> count()` taxa from bryophytes.
The number of leaves for the two gradients differed (reference = 7061, bird cliff = 4284).

```{r}
#| label: fig-traits
#| fig-height: 10
#| fig-cap: Trait distributions from six sites in Svalbard. Distributions of trait data (unweighted values) based on all sampled leaves (itex warming experiment and gradients) in Svalbard. The size traits (height, mass, length, area and thickness) are log transformed.

traits <- traits_itex_clean |> 
  # remove Betula nana leaves
  filter(!is.na(Site)) |> 
  mutate(location = case_when(Site == "DH" ~ "Dryas heath",
                              Site == "CH" ~ "Cassiope heath",
                              Site == "SB" ~ "Snowbed")) |> 
  bind_rows(traits_gradient_clean |> 
  mutate(location = if_else(Gradient == "C", "Reference", "Bird cliff"))) |> 
  #log transform size and area traits
      mutate(
        Value = if_else(
          Trait %in% c(
            "Plant_Height_cm",
            "Wet_Mass_g",
            "Dry_Mass_g",
            "Leaf_Area_cm2",
            "Leaf_Thickness_mm",
            "Shoot_Length_cm",
            "Shoot_Length_Green_cm"
          ),
          true = suppressWarnings(log(Value)),# suppress warnings from log(-value) in isotopes (these are calculated but not kept)
          false = Value
        )) |>
      # order traits
      mutate(Trait = factor(Trait, levels = c("Plant_Height_cm", "Dry_Mass_g", "Wet_Mass_g","Leaf_Area_cm2", "Leaf_Thickness_mm", "Shoot_Length_cm", "Shoot_Length_Green_cm", "Shoot_ratio", "SLA_cm2_g", "LDMC", "WHC_g_g", "SSL_cm_g", "C_percent", "N_percent", "CN_ratio", "P_percent", "NP_ratio", "dC13_permil", "dN15_permil")),
             location == factor(location, levels = c("Reference", "Bird cliff", "Dryas heath", "Cassiope heath", "Snowbed")))
  
vasc <- traits |> 
  filter(Functional_group == "vascular") |> 
  ggplot(aes(x = Value, fill = location)) +
  geom_density(alpha = 0.7) +
  scale_fill_viridis_d(option = "magma", begin = 0.2) +
  labs(x = "", tag = "a)") +
  facet_wrap(~Trait, scales = "free") +
  theme_bw()

bryo <- traits |> 
  filter(Functional_group == "bryophyte") |> 
  ggplot(aes(x = Value, fill = location)) +
  geom_density(alpha = 0.7) +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.7) +
  labs(x = "", tag = "b)") +
  facet_wrap(~Trait, scales = "free") +
  theme_bw()

vasc / bryo + plot_layout(guides = "collect")

```


Unweighted trait distributions of all trait data per site are provided in Fig. X. 
The trait data covered between 57% and 85% of the cumulative covers of the species in the plant community, based on the sum of covers from the control plots (calculations based on datasets i).
The leaf traits show consistent trends with elevation.
The “size-related traits” such as height, mass and area tend to decrease towards higher elevations (Fig. 4).
LDMC shows a decreasing trend, indicating more stress-tolerant leaves at higher elevations. SLA does not show a clear trend with elevation.


### Soil C and N

```{r}
n <- soil_gradient_clean |> 
  group_by(Gradient, Variable) |> 
  summarise(n = n())
  
model <- soil_gradient_clean |>
  mutate(Gradient = factor(Gradient, levels = c("C", "B"))) |> 
  group_by(Variable) |> 
  nest(data = -c(Variable)) |> 
  mutate(model = map(data, ~lme(Value ~ Gradient * Elevation_m, random = ~1|PlotID, data = .x)),
         result = map(model, tidy)) %>%
  unnest(result) |> 
  filter(effect == "fixed") |> 
  select(-data, -model, -effect, -group)
```

Vegetation height and structure data from the gradient has a total of `r nrow(soil_gradient_clean)` observations (gradient x site x plot x variable; Table 1, Table 8).
There are 21 observations for C and N each at the reference gradient and 14 each at the bird cliff.
Carbon and nitrogen content did not vary between the gradients or with elevation.

For an overview over the clean datasets and links to the code to clean and extract these data from the raw data, see Table 1. The final clean data is provided in the "Soil" folder, and the raw data is provided in the "RawData" folder on OSF(ref). The code to download and clean the data can be found in the GitHub repository(ref) in the file R/soil_plan.R.


### Climate

```{r}
climate_itex_clean |> 
  group_by(Variable) |> 
  summarise(mean = round(mean(Value), 2),
            se = round(sd(Value)/sqrt(n()), 2))
```

Climate data from the ITEX warming experiment has a total of `r nrow(climate_itex_clean)` observations including temperature, PAR, relative humidity, water content, and solar radiation data throughout 2015 - 2018 (date x variable; Table 1, Table X). 

```{r}
temperature_itex_clean |> 
  group_by(LoggerType, LoggerLocation) |> 
  summarise(mean = round(mean(Value), 2),
            se = round(sd(Value)/sqrt(n()), 2))
```

Temperature data from the ITEX warming experiment has a total of `r nrow(temperature_itex_clean)` observations from 2004 - 2005 and 2015 - 2018 (date x site x treatment x logger; Table 1, Table Y). 



```{r}

model <- climate_gradient_clean |>
  left_join(coords |> 
              mutate(Site = as.character(Site)), 
            by = c("Gradient", "Site", "PlotID")) |> 
  mutate(Gradient = factor(Gradient, levels = c("C", "B"))) |> 
  group_by(Variable) |> 
  nest(data = -c(Variable)) |> 
  mutate(model = map(data, ~lme(Value ~ Gradient * Elevation_m, random = ~1|PlotID, data = .x)),
         result = map(model, tidy)) %>%
  unnest(result) |> 
  filter(effect == "fixed") |> 
  select(-data, -model, -effect, -group) |> filter(p.value < 0.05)
```

Climate data from the gradient has a total of `r nrow(climate_gradient_clean)` observations (n = 81 for soil moisture and temperature each; gradient x site x plot x variable; Table 1, Table Z).
Soil temperature was higher at the reference gradient (E = 2.68, t71 = 6.11, P < 0.001). Soil moisture decreased with elevation but more strongly at the bird (E = -0.10, t71 = -3.42, P = 0.001).

For an overview over the clean datasets and links to the code to clean and extract these data from the raw data, see Table 1. The final clean data is provided in the "Climate" folder, and the raw data is provided in the "RawData" folder on OSF(ref). The code to download and clean the data can be found in the GitHub repository(ref) in the file R/climate_plan.R.
